---
title: "k-means clustering"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we'll run k-means clustering on the `pokemon` dataset from [Kaggle](https://www.kaggle.com/datasets/abcsds/pokemon?resource=download).

The number of clusters is not known beforehand, so testing different `center` parameters may be a good idea. 

```{r}
suppressPackageStartupMessages(library(dplyr))
set.seed(1000)

pokemon <- read.csv('pokemon.csv')

# subset features
df <- pokemon %>% select(HP, Attack, Defense, Sp..Atk, Sp..Def, Speed)

# Initialize total within sum of squares error: wss
wss <- 0

# Look over 1 to 15 possible clusters
for (i in 1:15) {
  # Fit the model: km.out
  km.out <- kmeans(df, centers = i, nstart = 20, iter.max = 50)
  # Save the within cluster sum of squares
  wss[i] <- km.out$tot.withinss
}
```

Now we create a scree plot of these kmeans clusters to determine the best number of centers

```{r}
# Produce a scree plot
plot(1:15, wss, type = "b", 
     xlab = "Number of Clusters", 
     ylab = "Within groups sum of squares")
```

We can choose 3 as the number of clusters

```{r}
# Build model with k clusters: km.out
km.out <- kmeans(df, centers = 3, nstart = 20, iter.max = 50)

# View the resulting model
km.out
```

Now we can examine cluster characteristics

```{r}
# create column for clusters
df$cluster <- km.out$cluster

# summarize data by group
df.summ <- df %>%
  group_by(cluster) %>%
  summarize(mean.HP = mean(HP),
            mean.Attack = mean(Attack),
            mean.Defense = mean(Defense),
            mean.Sp.Atk = mean(Sp..Atk),
            mean.Sp.Def = mean(Sp..Def),
            mean.Speed = mean(Speed))

print(df.summ)
```


We can see from the summary statistics that:

* Cluster `r unlist(df.summ[which.max(df.summ$mean.HP),1])` has the highest mean HP
* Cluster `r unlist(df.summ[which.max(df.summ$mean.Attack),1])` has the highest mean attack
* Cluster `r unlist(df.summ[which.max(df.summ$mean.Defense),1])` has the highest mean defense
* Cluster `r unlist(df.summ[which.max(df.summ$mean.Sp.Atk),1])` has the highest mean special attack
* Cluster `r unlist(df.summ[which.max(df.summ$mean.Sp.Def),1])` has the highest mean special defense
* Cluster `r unlist(df.summ[which.max(df.summ$mean.Speed),1])` has the highest mean speed
