---
title: "Principal Component Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

First we'll run k-means clustering on the `pokemon` dataset from [Kaggle](https://www.kaggle.com/datasets/abcsds/pokemon?resource=download).

```{r}
suppressPackageStartupMessages(library(dplyr))
set.seed(1000)

pokemon <- read.csv('pokemon.csv')

# subset features
df <- pokemon %>% select(HP, Attack, Defense, Sp..Atk, Sp..Def, Speed)
```

## Execute PCA

Then we'll run a principle component analysis on the data

```{r}
pr.out <- prcomp(df, center = TRUE, scale = TRUE)
summary(pr.out)
```

## Model

To examine the variance explained, we can extract the standard deviation from the model and square it

```{r}
pr.var <- pr.out$sdev^2
pve <- pr.var / sum(pr.var)
```

And build a scree plot for both proportion of variance explained and cumulative proportion of variance explained

```{r}
plot(pve, xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     ylim = c(0, 1), type = "b")

plot(cumsum(pve), xlab = "Principal Component",
     ylab = "Cumulative Proportion of Variance Explained",
     ylim = c(0, 1), type = "b")
```

The plot shows that it takes 3 features to explain 75% or more of the variance

## Contribution of each variable

To determine how much each variable is represented in a given component, we can use the Cos2 plot (square cosine)

```{r}
library(factoextra)
fviz_cos2(pr.out, choice = "var", axes = 1:2)
```

This computes the square cosine value for each variable with respect to the first two principal components. The most represented variables are speed and defense.